Automated Amortized Resource Bound Analysis \\ for Control Flow Graphs

### 2015-09-06 cfg/main.tex
cfg rules

% Comment for anonymization
% Add preprint option for the extended version
% Also look for "SHORT" comments
 \def\fullversion{}

\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}

% Structure:
%
%    Abstract                              1c
% 1) Intro                                 3c
% 2) The Potential Method                  1c
% 3) Compositional Resource-Bound Analysis 4c
%    - Gulwani/ examples
% 4) Derivation Rules                      2c
% 5) Automatic Inference via LP Solving    2c
% 6) Logical State and User Interaction    2c
% 7) Soundness Proof                       2c
%    1) Cost Semantics for Clight
%    2) Quantitative Hoare Logic
% 8) Experimental Evaluation               2c
% 9) Related Work                          1c
% a) Conclusion                            1c

\newcommand{\iffull}[2]{\ifx\fullversion\undefined{#2}\else{#1}\fi}
\newcommand{\ifshort}[2]{\ifx\fullversion\undefined{#1}\else{#2}\fi}

\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
\newcommand{\paraskip}[0]{\ifshort{\vspace{-4pt}}{}}
\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\captionsetup[table]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}
\usepackage{enumitem}
\usepackage{array} %for stupid tables
\usepackage{hhline}

%----------------------------------------------------

\usepackage{prettyref}
\newcommand{\pref}[1]{\prettyref{#1}}
\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
\newcommand{\ppref}[1]{\vpageref[]{#1}}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{app}{Appendix~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{cha}{Challenge~\ref{#1}}
\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}

%----------------------------------------------------

\usepackage{amsthm}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

%----------------------------------------------------

\input{definitions}
\begin{document}

\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}

%SHORT
\ifx\fullversion\undefined
\abovedisplayskip=3pt
\belowdisplayskip=3pt
\abovedisplayshortskip=3pt
\belowdisplayshortskip=3pt
\else
\fi
%SHORT

\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
\copyrightyear{2014}
\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}

%----------------------------------------------------

\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}

\titlebanner{\iffull{Extended version}{}}%{Draft -- Do not distribute}        % These are ignored unless
\preprintfooter{\iffull{YALEU/DCS/TR-1505\hspace{-.8cm}}{Draft\hspace{.5cm}}}%{Draft}   % 'preprint' option specified.

\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

\authorinfo
{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
{Yale University \vspace{-1cm}}

@@

\begin{abstract}
  This paper presents a new approach for automatically deriving
  worst-case resource bounds for C programs.
  %
  The described technique combines ideas from amortized analysis and
  abstract interpretation in a unified framework to address four
  challenges for state-of-the-art techniques: compositionality, user
  interaction, generation of proof certificates, and scalability.
  %
  \emph{Compositionality} is achieved by incorporating the potential
  method of amortized analysis.  It enables the derivation of global
  whole-program bounds with local derivation rules by naturally
  tracking size changes of variables in sequenced loops and function
  calls.  The resource consumption of functions is described abstractly
  and a function call can be analyzed without access to the function
  body.
  %\
  \emph{User interaction} is supported with a new mechanism that
  clearly separates qualitative and quantitative verification. A user
  can guide the analysis to derive complex non-linear bounds by using
  auxiliary variables and assertions.  The assertions are
  separately proved using established qualitative techniques such as
  abstract interpretation or Hoare logic.
  %
  \emph{Proof certificates} are automatically generated from the local
  derivation rules. A soundness proof of the derivation system with
  respect to a formal cost semantics guarantees the validity of the
  certificates.
  %
  \emph{Scalability} is attained by an efficient reduction of
  bound inference to a linear optimization problem that can be solved
  by off-the-shelf LP solvers.
  %
  The analysis framework is implemented in the publicly-available
  tool $\toolname$. An experimental evaluation demonstrates the
  advantages of the new technique with a comparison of $\toolname$ with
  existing tools on challenging micro benchmarks and the analysis of
  more than $\totalLoc$ lines of C code from the cBench benchmark suite.

% Contrary to most available tools for
% C, ours has a compositional design that provides the
% programmer with bounds expressed in terms of meaningful
% variables.  The compositionality allows to handle uniformly
% nested sequenced loops and recursive sequenced function
% calls.  The automatic analysis is implemented
% as a tool that leverages state-of-the-art SMT solvers
% LP solvers to automatically infer resource bounds on
% a wide range of C programs.  It is the first automatic
% amortized analysis that can derive bounds that depend
% on negative integers and differences between numbers.
% To prove its soundness and give a concrete shape to proof
% certificates, we developed a quantitative Hoare logic
% that we proved sound using the Coq proof assistant.  This
% logic allows manual proofs of arbitrarily complex resource
% bounds on C programs and interaction with the automation.
% To prove the scalability and precision of this new tool,
% we run it on multiple code snippets from the literature,
% reproduce results from previous work, and analyze a
% consequent part of a modern benchmark suite.

%   The goal of quantitative resource analysis is to provide developers
%   with quantitative information about the runtime behavior of software
%   at development time.  Recent years have seen tremendous progress in
%   automatically deriving worst-case resource bounds, yet many
%   challenges in specifying, interactively deriving, formally
%   certifying, and composing resource bounds remain unsolved.
%
%   This paper describes a novel quantitative resource analysis
%   framework that tackles these challenges for C programs.  The
%   analysis framework consists of three parts: First, a parametric cost
%   semantics formalizes the resource consumption of CompCert Clight
%   programs.  Second, a quantitative Hoare logic enables users to
%   interactively develop resource bounds in the Coq Proof Assistant.
%   The logic is proved sound with respect to the cost semantics, and
%   shallow embedding enables a derived bound to be any function that is
%   definable in Coq.  Third, an automatic amortized resource analysis
%   for Clight computes derivations in the quantitative Hoare logic.
%   It is the first automatic amortized analysis that can derive bounds
%   that depend on negative integers or differences between numbers,
%   which is crucial to handle typical systems code.  Both, the
%   quantitative logic and the automatic amortized analysis are
%   naturally compositional and can be combined to semi-automatically
%   derive global resource bounds.
%   %
%   An experimental evaluation demonstrates the practicality of the
%   analysis framework.  The expressivity of the logic is shown by
%   manually deriving customized bounds that are tailored to specific
%   algorithms.  A comparison of the automatic amortized analysis with
%   other automatic tools on 30 challenging loop and recursion patterns
%   from the literature and open-source software shows that the bounds
%   derived by the automatic amortized analysis are often more precise.
\end{abstract}

% \category{D.2.4}{Software Engineering}
% {Software/Program Verification}
% \category{F.3.1}{Logics and Meanings of Programs}
% {Specifying and Verifying and Reasoning about Programs}

% \terms Verification, Reliability

% \keywords Formal Verification, Compiler Construction, Program Logics,
% Stack-Space Bounds, Quantitative Verification

\sectskip
\section{Introduction}
\label{sec:intro}
\aftersectskip

In software engineering and software verification,
we often would like to have static information
about the quantitative behavior of programs.
For example, stack and heap-space bounds
are important to ensure the reliability of
safety-critical systems~\cite{Regehr05}.
Static energy usage information is critical
for autonomous systems and has applications in
cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help create
constant-time implementations that prevent
side-channel attacks~\cite{KasperS09,BartheBCLP14}.
Loop and recursion-depth bounds are used to
ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and
complexity bounds are needed to verify cryptographic
protocols~\cite{BartheGB09}.  In general, quantitative
resource information % at design time
can provide useful
feedback for developers.

% NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW NEW

Available techniques for {\em automatically} deriving worst-case
resource bounds fall into two categories.  Techniques in the first
category derive impressive bounds for numerical imperative programs,
but are not compositional.  This is problematic if one needs to derive
global whole-program bounds.
%%%%%%
Techniques in the second category derive
tight whole-program bounds for programs with regular loop or
recursion patterns that decrease the size of an individual variable or
data structure. They are highly compositional, scale for large
programs, and work directly on the syntax.  However,
they do not support multivariate interval-based resource bounds
(e.g., $x-y$) which are common in C programs. Indeed,
it has been a long-time open problem to develop compositional
resource analysis techniques that can work for typical imperative code
with non-regular iteration patterns, signed integers, mutation, and non-linear
control flow. % As a result, they are not suitable for
% typical imperative code.

Tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least two ways.
First, they all base their analysis on some form of
\emph{ranking function} or \emph{counter instrumentation}
that is linked to a local analysis.  As a result, loop bounds are
arithmetic expressions that depend on the values of
variables just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and function calls in
terms of the input parameters of a function.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract resource behavior;
instead, they have to inline the procedure body to perform their
analysis.

Tools in the second category originate form the {\em potential method}
of amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object-oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs.
However, a major weakness of existing potential-based techniques
is that they can only associate \emph{potential}
with individual program variables or data structures.
For C programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y-i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

% these
% techniques fail for numerical imperative programs because they lack
% fine-grained data-structure information to hook the analysis on.
% Another issue is that existing techniques associate \emph{potential}
% directly with single program variables.

A general problem with existing tools (in both categories) is user interaction. When a tool
fails to find a resource bound for a program, there is no possibility
for sound user interaction to guide the tool during bound
derivation.
For example, there is no concept of manual proofs of
resource bounds; and no framework can support composition
of manually derived bounds with automatically inferred bounds.

% Finally, the tools often operate on an abstract version of the and use
% complex external tools from term rewriting and abstract
% interpretation. This makes it hard to generate a proof certificate for
% the bounds or even to prove the soundness of the tools with respect to
% a formal cost semantics.

This paper presents a new compositional framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two aforementioned
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis--based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a functional specification that can be used at every
call site without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that may
become available during execution (e.g., when freeing
memory).  Compared to more classical
approaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovations that make amortized analysis work
on imperative languages are to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can be solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

The use of the potential method of amortized analysis makes user
interaction possible in different ways. For one thing, we can directly
combine the new automatic analysis with manually derived bounds in a
previously-developed quantitative Hoare logic~\cite{veristack14} (see
\pref{sec:soundness}).  For another thing, we describe a new mechanism that
allows the separation of quantitative and qualitative verification
(see \pref{sec:anno}).  Using this mechanism, the user can guide the
analysis by using auxiliary variables and logical assertions that can
be verified by existing qualitative tools such as Hoare logic or
abstract interpretation.  In this way, we can benefit from existing
automation techniques and provide a middle-ground between fully
automatic and fully manual verification for bound derivation. This
enables the semi-automatic inference of non-linear bounds, such as
polynomial, logarithmic, and exponential bounds.

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than $\totalLoc$ lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but one of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

Examples from cBench and micro benchmarks demonstrate the practicality and
expressiveness of the user guided bound inference.  For example, we
derive a logarithmic bound for a binary search function\iffull{, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, }{ }and a bound
that amortizes the cost of $k$\iffull{ successive }{ }increments to a binary
counter (see \pref{sec:anno}).

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that may
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than $\totalLoc$ lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

\sectskip
\section{The Potential Method}
\aftersectskip

The idea that underlies the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes
on a starting state $\state$ and consumes $n$ resource units of
some user-defined quantity.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after
the execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a program
state such that $(S, \state) \Downarrow_n \state'$.  Then
$\Phi(\state)$ is a valid resource bound.

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by focusing on the sequencing of
programs.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

% \subsection{Automating  Amortized Analysis}

% To make amortized reasoning feasible on large code bases, we would
% like to automatically infer the potential functions.  The first step
% towards this goal is to reduce our search space by fixing the shape of
% potential functions $\Phi$.  In this paper, we focus on a form that
% can express tight bounds for many typical programs and allows
% inference view \emph{linear programming}:
% %
% $$
% \Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
%   q_{(x,y)} \cdot |[\state(x),\state(y)]|
% $$
% where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
% that a program state is a simple map from program variables to integers
% and we treat constants as program variables (that cannot be changed
% by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
% the sum.  % We then develop an inference rule for each syntactic construct
% % that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
% % the previous section.

The derivation of a resource bound using potential functions is best
explained by example.  If we use a cost model that assigns cost $n$ to
the function call \code{tick(n)} and cost $0$ to all other operations
then the cost of the following example can be bounded by
$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
  % the signs of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound, we start with the initial potential $\Phi_0 =
|[x,y]|$, which
we also use as the loop invariant.  For the loop body we have (like in
Hoare logic) to derive a triple
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential
$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
we denote the updated version of $x$ after the assignment by $x'$ then
the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
after the assignment \code{x=x+1} holds.  This means that we have the
potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
\code{tick(1)} consumes one resource unit, we end up with potential
$|[x,y]|$ after the loop body and have established the loop invariant
again.

\begin{figure}[t]
  \centering
\vspace{-1.5ex}
\begin{lstlisting}[mathescape]
${\color{blue}%
  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
while (x+K<=y) {
  ${\color{blue}%
  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
  x=x+K;
  ${\color{blue}%
  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| \} }$
  tick(T);
  ${\color{blue}%
  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
}
${\color{blue}%
 \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| \} }$
\end{lstlisting}
  \caption{Derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote concrete constants.
    % The reasoning works uniformly for all such $K$ and
    % $T$.
  }
  \label{fig:ex1}
\end{figure}

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
logical assertions from potential functions with semicolons. Note that the
logical assertions are only used in the rule for the assignment
\code{x=x+K}.

To the best of our knowledge, no other implemented tool for C is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by
hand. We will show in the following sections that the
potential method makes automatic bound derivation straightforward.

% To automate the reasoning, we first introduce an unknown rational
% variable for each factor in the potential functions.  We then use our
% inference rules (see \pref{sec:AAA}) to emit linear constraints on
% these variables that enforce that variable assignments that respect
% the constraints correspond to sound potential annotations.  For
% instance if $K = 1$ in \pref{fig:ex1} then we would have the
% annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
% |[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
% p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
% are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
% $p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
% q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
% the coefficients of the initial potential function. LP solvers are
% among the most mature constraint solving tools. As a result, the bound
% derivation works very efficiently in practice.

% \iffull{It might be surprising that we}{We} only track simple linear relations in
% the logical context $\Gamma$.  While it would of course be possible to keep
% track of more sophisticated assertions, this simple form is
% sufficient for the examples we considered and we can efficiently
% decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
% assignment rule.  Nevertheless, this logical part of our analysis
% system is orthogonal to the quantitative part and can be easily
% extended if necessary.

The concept of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#

#\!\!\!\!\!#while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.

\iffull{In general, the potential based approach addresses several challenges
in resource bound analysis. This includes amortized bounds in which
individual loop iterations have different resource consumption,
sequential loops or function calls that require the tracking of size
changes of variables, and nested loops that interact by modifying the
same set of variables.}{}

\newlength{\progwidth}

\begin{figure*}
 \setlength{\progwidth}{.22\linewidth}
  \centering
\vspace{-.3cm}
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
  if (m>y)
    ${\color{blue}%
    \{ m{>}y; \, |[x, n]| {+} |[y, m]| \} }$
    y=y+1;
    ${\color{blue}%
    \{ \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  else
    ${\color{blue}%
    \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
    x=x+1;
    ${\color{blue}%
    \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  ${\color{blue}%
  \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  tick(1);
} ${\color{blue}%
  \{ \cdot; \, |[x, n]| {+} |[y, m]| \} }$
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

>>>>

\documentclass[letterpaper,11pt]{article}

% \documentclass[preprint]{sigplanconf}
\usepackage[numbers]{natbib}

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\usepackage{mathpartir} %for typing rules
%\usepackage{graphicx} %to include png images
\usepackage{microtype} %better micro typing
\usepackage[standard,amsmath,thmmarks]{ntheorem}
\usepackage{mathabx} %for blackdiamond

\usepackage{prettyref}
\newcommand{\pref}[1]{\prettyref{#1}}
\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
\newcommand{\ppref}[1]{\vpageref[]{#1}}

\input{definitions}

%------------Draft---------------------------------------
\usepackage{eso-pic}
\usepackage{xcolor}
\AddToShipoutPicture{\setlength{\unitlength}{1in}\put(4.25,10.0){\makebox[0pt][c]{\normalfont\textsf{\textbf{\large \color{gray} Draft -- \today}}}}}
%--------------------------------------------------------

\newcommand{\jan}[1]{{\color{blue}{[Jan: #1]}}}

\title{Automated Amortized Resource Bound Analysis \\ for Control Flow Graphs}

\author{Quentin Carbonneaux \and Jan Hoffmann}

\begin{document}

@@

\section{Control Flow Graphs}

\paragraph{Straight Line Statements}

Straight line statements (also just statements) are inductively
defined by the following grammar.
%
\begin{align*}
S &:= \code{tick}(n) \mid \code{assert}~E
\mid \code{skip}
\mid x \gets E
\mid x \gets f(x^*)
\mid S;S
\end{align*}
%
Expressions $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free. We write $\statements$ for the set of statements.

\paragraph{Control Flow Graphs}

A control flow graph (CFG) is a set a graph $G=(V,E,\labels)$ with
statement-labeled nodes. Here, $V$ is a set of nodes,
$E \subseteq V \times V$ is a set of edges, and
$\labels : V \to \statements$ is a mapping from nodes to statements.

\section{Operational Semantics of CFGs}

\jan{Here we probably need some non-deterministic choice on the
  edges. We have to watch out that this is consistent
  with the assertions that we insert in the blocks. We might have to
  put assertion on the edges to avoid accounting cost for a jump that
  might not happen.}

\section{Potential Annotations for CFGs}

\subsection{Linear Potential Functions.}

To find resource bounds automatically, we first need to restrict our
search space.  In this work, we focus on the following form of
potential functions, which can express tight bounds for many typical
programs and allows for inference with \emph{linear programming}.
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Q$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as 
$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$
%
We often write $xy$ to denote the
index $(x,y)$.  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of rational numbers where only
a finite number of elements are not zero.
